---
title: "PCA Examples"
author: "JDRS"
date: "January, 2020"
output:
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

###  Examples of PCA in R

Handy libraries

*   Nice graphical representations of correlations
*   Learn more [here](https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html)


```{r}
library(corrplot)
library(PerformanceAnalytics)
```


###  World Bank Data 2016


```{r}
#read in data from online location
masterdata
names(masterdata)

#Get subset of data that is complete for relevant variables
MSB <- masterdata[ ,c("Species", "DBH", "SoilGWC", "pH", "SIR", "Aspect", "Elevation", "Slope", "meanSoilT", "meanLitT", "meanSoilM")]
MSB <- masterdata[masterdata$Site == "SCBI" ,c("Species", "DBH", "SoilGWC", "pH", "SIR", "Aspect", "Elevation", "Slope", "meanSoilT", "meanLitT", "meanSoilM")]

#We need complete cases for code to work
MSB <- MSB[complete.cases(MSB),]
dim(MSB)

```


For starters, make correlation matrices and use visual representations.
```{r}
#make correlation matrix to see if PCA will work well - remove column of country names
round(cor(MSB[, -1]), 2)

#Cooler visual representation of correlations
corrplot(cor(MSB[, -1]), method = "ellipse")

#Cooler visual representation of correlations
#Order option orders variables based on data order, alphabetically, results of cluster analysis, etc.
#  See help file or link above to get details.

corrplot(cor(MSB[,-1]),method = "ellipse", order="FPC")
corrplot(cor(MSB[,-1]),method = "ellipse", order="AOE")
corrplot(cor(MSB[,-1]),method = "ellipse", order="hclust")

#Lots of options here - but my personal favorite
corrplot.mixed(cor(MSB[,-1]), lower.col = "black", upper = "ellipse", tl.col = "black", number.cex = .7, order = "hclust", tl.pos = "lt", tl.cex = .7)
```


It's also important to check for linearity.
```{r}
#make matrix plot to check for linearity
plot(MSB[,-1], pch=19, cex=.7, col='red', main="Matrix plot of WB raw data")

#Here is a cool way to look for non-linearity, get correlation, make histograms all at once.
chart.Correlation(MSB[, -1], histogram=TRUE, pch=19)

```

Let's see if data happens to have multivariate normal distribution (not required but handy for parallel analysis)

```{r}
#get online function
source("http://www.reuningscherer.net/STAT660/R/CSQPlot.r.txt")

#run the function
CSQPlot(MSB[,-1],label="MSB SCBI Data")
```

Seems like transformed variables are closer to multivariate normal.  Let's get this data for reference.

```{r}
WBtrans <- WB[,c("Country","Rural","logGNI","logImports","logExports","Cell","Fertility16","InfMort","LifeExp","PM2.5","logCO2","Diesel","logEnergyUse")]
WBtrans <- WBtrans[complete.cases(WBtrans),]

#run the function
CSQPlot(WBtrans[,-1],label="World Bank Data")

```


Here is correlation plot for transformed data

```{r}
corrplot.mixed(cor(WBtrans[,-1]), lower.col="black", upper = "ellipse", tl.col = "black", number.cex=.7, order = "hclust",
               tl.pos = "lt", tl.cex=.7, main="Correlations for Transformed WB Data")
```

Check linearity with matrix plots

```{r}
#make matrix plot to check for linearity
plot(WBtrans[,-1], pch=19, cex=.7, col='red', main="Matrix plot of WB transformed data")

#Here is a cool way to look for non-linearity, get correlation, make histograms all at once.
chart.Correlation(WBtrans[, -1], histogram=TRUE, pch=19)
```






####  RUN PCA ON RAW WB DATA

NOTE : there are multiple funcions and packages that will perform PCA in R.

There is a nice discussion [here](http://www.gastonsanchez.com/visually-enforced/how-to/2012/06/17/PCA-in-R/ )  

Included in R are princomp() and prcomp()   (very similar)

#####FIRST, use princomp()

```{r}
pc1 <- princomp(MSB[,-1], cor = TRUE)
```


Objects created : loadings are the eigenvectors, scores are, well, the scores, sdev is sqrt of eigenvalues!

```{r}
names(pc1)

#print results
print(summary(pc1), digits = 2, loadings = pc1$loadings, cutoff=0)
```


Get eigenvalues - NOTE that R stores the standard deviation for each PC, i.e. it has the square root of the eigenvalues!

```{r}
round(pc1$sdev^2,2)
```

Make a screeplot  


```{r}
screeplot(pc1,type="lines",col="red",lwd=2,pch=19,cex=1.2,main="Scree Plot of Raw WB Data")
```

Perform parallel analysis.  The code below includes a

```{r}
#get the function online
source("http://www.reuningscherer.net/STAT660/R/parallel.r.txt")

#make the parallel analysis plot using the parallelplot function
parallelplot(pc1)
```

Make scoreplot with confidence ellipse as well as a biplot.

```{r}
#  c(1,2) specifies to use components 1 and 2
#get function from online
source("http://reuningscherer.net/stat660/r/ciscoreplot.R.txt")

#run the function
ciscoreplot(pc1,c(1,2),MSB[,1])

#make a biplot for first two components
biplot(pc1,choices=c(1,2),pc.biplot=T)


```


####SECOND, use PCA() function in FactoMineR on RAW World Bank Data

You can learn about FactoMineR in this [video]?( https://www.youtube.com/watch?v=CTSbxU6KLbM&list=PLnZgp6epRBbTsZEFXi_p6W48HhNyqwxIu&index=3)

```{r}
library(FactoMineR)

#Make rownames of MSB equal to country names
rownames(MSB) <- MSB[,1]

#The PCA function scales variables by default, and gives a scoreplot and a 'correlation' biplot.  Notice results are flipped vertically from results above.

par(cex=.8, col="blue")
pc2 <- PCA(MSB[,-1])
dev.off()
summary(pc2)

#make a screeplot - oddly, not easy in this package . . .  
#parallel analysis,etc, use princomp() or prcomp()

```

####  RUN PCA ON TRANSFORMED DATA


```{r}
pc1_trans <- princomp(WBtrans[,-1], cor=TRUE)

#print results
print(summary(pc1_trans),digits=2,loadings=pc1_trans$loadings,cutoff=0)

#Get eigenvalues
round(pc1_trans$sdev^2,2)

#make a screeplot  
screeplot(pc1_trans,type="lines",col="red",lwd=2,pch=19,cex=1.2,main="Scree Plot of Transformed WB Data")

#perform parallel analysis
#get the function online
source("http://www.reuningscherer.net/STAT660/R/parallel.r.txt")

#make the parallel analysis plot
parallelplot(pc1_trans)

#make scoreplot with confidence ellipse : 
#  c(1,2) specifies to use components 1 and 2
#get function from online
source("http://reuningscherer.net/stat660/r/ciscoreplot.R.txt")

#run the function
ciscoreplot(pc1_trans,c(1,2),WBtrans[,1])

#make a biplot for first two components
biplot(pc1_trans,choices=c(1,2),pc.biplot=T)
```


##   Environmental Attitudes Data

```{r}
#get data : change path to wherever you have data.
envatt <- read.delim("http://reuningscherer.net/stat660/data/EnvAttitudes.1974.txt",header=T,sep="\t")

#Keep complete cases
envatt <- na.omit(envatt)

#Notice that a matrix plot to check for linearity is useless here
plot(envatt, pch=19, cex=.7, col='red', main="Matrix plot of Env Attitude Data")

#Here is correlation plot - correlations are a bit weak!
corrplot.mixed(cor(envatt), lower.col="black", upper = "ellipse", tl.col = "black", number.cex=.7, order = "hclust",
               tl.pos = "lt", tl.cex=.7, main="Correlations for Env Attitude Data")

```

Perform PCA

```{r}
#perform PCA
pc3<-princomp(envatt[complete.cases(envatt),],cor=TRUE)

#print results
print(summary(pc3),digits=2,loadings=pc3$loadings,cutoff=0)

#view eigenvalues (note that R gives square-root of eigenvalues)
round(pc3$sdev^2,2)

#make a scree plot with parallel analysis
source("http://www.reuningscherer.net/STAT660/R/parallel.r.txt")
parallelplot(pc3)


#make scoreplot with confidence ellipse : 
#  c(1,2) specifies to use components 1 and 2
#get function from online
source("http://reuningscherer.net/stat660/r/ciscoreplot.R.txt")

#run the function
ciscoreplot(pc3,c(1,2),c(1:pc3$n.obs))


#make a biplot for first two components

biplot(pc3,choices=c(1,2),pc.biplot=T, cex=.7)
```

###  PCA - NASA Understory Data

```{r}
#read in data from online location
NASA <- read.csv("http://reuningscherer.net/stat660/data/NASA1/snf/NASAUnderstory.csv",header=T)
head(NASA)
names(NASA)

#get long version of species columns labels
NASA.lab <- NASA$Labels
NASA <- NASA[,-32]

head(NASA)

#Correlation Plot
corrplot.mixed(cor(NASA[,-c(1:2)]), lower.col="black", upper = "ellipse", tl.col = "black", number.cex=.7, order = "hclust",
               tl.pos = "lt", tl.cex=.7, main="Correlations for NASA Data")

#Run PCA - remove column ID and species name
pc4 <- princomp(NASA[,-c(1:2)],cor=TRUE)

#print results
print(summary(pc4),digits=2,loadings=pc4$loadings,cutoff=0)

#view eigenvalues (note that R gives square-root of eigenvalues)
round(pc4$sdev^2,2)
```

```{r}

#Evaluate Multivariate Normality
#get online function
source("http://www.reuningscherer.net/STAT660/R/CSQPlot.r.txt")

#run the function
CSQPlot(NASA[,-c(1:2)],label="NASA Understory")

#make a scree plot with parallel analysis
source("http://www.reuningscherer.net/STAT660/R/parallel.r.txt")
parallelplot(pc4)


#make scoreplot with confidence ellipse: 
#  c(1,2) specifies to use components 1 and 2
#get function from online
source("http://reuningscherer.net/stat660/r/ciscoreplot.R.txt")

#run the function
ciscoreplot(pc4, c(1,2), c(1:pc4$n.obs))

#I was curious which overstory species was on each side
text(pc4$scores[,1],pc4$scores[,2], labels = NASA[,2], cex = 0.6, col = as.numeric(NASA[,2]))


#make a biplot for first two components
biplot(pc4, choices=c(1,2), pc.biplot = T, cex = 0.7)

```


